---
title: Linear equations, inequalities, sets and functions, quadratics, and logarithms
---

```{r common-r, code = xfun::read_utf8(here::here("_common.R")), include = FALSE}
```

```{r setup}
library(tidyverse)
```

# Simplify expressions

Simplify the following expressions as much as possible:^[Gill 1.1]

a. $(-x^4y^2)^2$

    <!--
    
    1. Distribute exponents over products.
    
        $$(-1)^2 x^{(2 \times 4)} y^{(2 \times 2)}$$
        
    1. Multiply 2 and 2 together.
    
        $$(-1)^2 x^{(2 \times 4)} y^4$$
        
    1. Multiply 2 and 4 together.
    
        $$(-1)^2 x^8 y^4$$
        
    1. Evaluate $(-1)^2$.
    
        $$x^8 y^4$$
        
    -->
        
a. $9(3^0)$

    <!--

    1. Any nonzero number to the zero power is 1.
    
        $$9(1)$$
        
    1. Anything times 1 is the same value.
    
        $$9$$

    -->
    
a. $(2a^2)(4a^4)$

    <!--

    1. Combine products of like terms.
    
        $$2 a^2 \times 4 a^4 = 2  \times 4a^{(2 + 4)}$$
        
    1. Evaluate $2 + 4$.
    
        $$2 \times 4 a^6$$
        
    1. Multiply 2 and 4 together.
    
        $$8 a^6$$
        
    -->
    
a. $\dfrac{x^4}{x^3}$

    <!--

    1. For all exponents, $\dfrac{a^n}{a^m} = a^{(n - m)}$.
    
        $$x^{(4 - 3)}$$
        
    1. Evaluate $4 - 3$.
    
        $$x$$
        
    -->
    
a. $(-2)^{7-4}$

    <!--

    1. Subtract $4$ from $7$.
    
        $$(-2)^3$$
        
    1. In order to evaluate $2^3$ express $2^3$ as $2 \times 2^2$.
    
        $$-2 \times 2^2$$
        
    1. Evaluate $2^2$.
    
        $$-2 \times 4$$
        
    1. Multiply $-2$ and $4$ together.
    
        $$-8$$
        
    -->
    
a. $\left(\dfrac{1}{27b^3}\right)^{1/3}$

    <!--

    1. Separate component terms.
    
        $$\dfrac{1}{27}^{1/3} \times \dfrac{1}{b^3}^{1/3}$$
        
    1. Evaluate cube roots.
    
        $$\dfrac{1}{3} \times \dfrac{1}{b}$$
        
    1. Combine terms.
    
        $$\dfrac{1}{3b}$$
    
    -->
    
a. $y^7y^6y^5y^4$

    <!--

    1. Combine products of like terms.
    
        $$y^{(7 + 6 + 5 + 4)}$$
        
    1. Evaluate $7 + 6 + 5 + 4$.
    
        $$y^{22}$$
        
    -->
    
a. $\dfrac{2a/7b}{11b/5a}$

    <!--

    1. Write as a single fraction by multiplying the numerator by the reciprocal of the denominator.
    
        $$\dfrac{2a}{7b} \times \dfrac{5a}{11b}$$
        
    1. Product property of exponents: $x^a \times x^b = x^{(a+b)}$
    
        $$\dfrac{5a \times 2a}{7b \times 11b} = \dfrac{5 \times 2a^{1+1}}{7 \times 11b^{1+1}}$$
        
    1. Evaluate $1+1$.
    
        $$\dfrac{5 \times 2a^2}{7 \times 11b^2}$$
        
    1. Multiple scalars together.
    
        $$\dfrac{10a^2}{77b^2}$$

    -->
    
a. $(z^2)^4$

    <!--

    1. Nested exponents rule: $(x^a)^b = x^{ab}$
    
        $$z^{2 \times 4}$$
        
    1. Evaluate $2 \times 4$
    
        $$z^8$$

    -->
    
# Simplify a (more complex) expression
 
Simplify the following expression:^[Gill 1.2]

$$(a+b)^2 + (a-b)^2 + 2(a+b)(a-b) - 3a^2$$
<!--

1. Expand $(a + b)^2$ with FOIL.

    $$a^2 + 2 a b + b^2 + (a - b)^2 + 2 (a + b) (a - b) - 3 a^2$$
    
1. Expand $(a - b)^2$ with FOIL.

    $$a^2 + 2 a b + b^2 + a^2 - 2 a b + b^2 + 2 (a + b) (a - b) - 3 a^2$$
    
1. Multiply $a + b$ and $a - b$ together using FOIL.

    $$a^2 + 2 a b + b^2 + a^2 - 2 a b + b^2 + 2 (a^2 - b^2) - 3 a^2$$
    
1. Distribute $2$ over $a^2 - b^2$.

    $$a^2 + 2 a b + b^2 + a^2 - 2 a b + b^2 + 2 a^2 - 2 b^2 - 3 a^2$$
    
1. Group like terms.

    $$(a^2 + a^2 + 2 a^2 - 3 a^2) + (b^2 + b^2 - 2 b^2) + (2 a b - 2 a b)$$
    
1. Combine like terms.

    $$a^2 + (b^2 + b^2 - 2 b^2) + (2 a b - 2 a b)$$
    
1. Look for the difference of two identical terms.

    $$a^2$$

-->

# Graph sketching

Let the functions $f(x)$ and $g(x)$ be defined for all $x \in \Re$ by
    
$$
f(x) = \left\{
    \begin{array}{ll}
        | x | & \quad \mbox{if} \ x < 1 \\
        1 & \quad \mbox{if} \ x \geq 1
    \end{array}
\right., \quad
g(x) = \left\{
    \begin{array}{ll}
        x^2 & \quad \mbox{if} \ x < 2 \\
        4 & \quad \mbox{if} \ x \geq 2
    \end{array}
\right.
$$

Sketch the graphs of:^[Pemberton and Rau problem 3-1]

1. $y = f(x)$
    
    <!--
    
    ```{r y-f-x, dependson = "base-plot"}
f_x <- function(x) {
  ifelse(x < 1, abs(x), 1)
}

function_plot +
  stat_function(fun = f_x, n = 1000) +
  ggtitle(expression(f(x)))
    ```
    
    -->
    
1. $y = g(x)$

    <!--

    ```{r y-g-x, dependson = "base-plot"}
g_x <- function(x) {
  ifelse(x < 2, x^2, 4)
}

function_plot +
  stat_function(fun = g_x, n = 1000) +
  ggtitle(expression(g(x)))
    ```
    
    -->

1. $y = f(g(x))$

    <!--

    To sketch the composite function, we first evaluate $g(x)$ for different values of $x$, and then evaluate $f(g(x))$ for different outputs of $g(x)$.
    
    * For $x \geq 2$, $g(x)$ is a constant value: 
        $$
        \begin{aligned}
        x \geq 2 \\
        g(x) = 4 \\
        f(g(x)) = f(4) = 1
        \end{aligned}
        $$
    
    * For $x < 2$, $g(x)$ is not constant: $g(x) = x^2$. $f(x)$ evaluates differently depending on its input, so we have two cases based on the output of $g(x)$:
    
        * if $g(x) < 1$, $f(g(x)) = |g(x)| = |x^2| = x^2$. This is the case when:
            $$
            \begin{aligned}
            g(x) < 1 \\
            x^2 <1 \text{ and } x < 2\\
            -1 < x < 1
            \end{aligned}
            $$
    
        * if $g(x) \geq 1$, $f(g(x)) = 1$. This is the case when:
            $$
            \begin{aligned}
            g(x) \geq 1 \\
            x^2 \geq 1 \text{ and } x < 2\\
            x \leq -1 \mbox{ or }  1 \leq x < 2
            \end{aligned}
            $$
    
    * Therefore, $f(g(x))$ has the following values:
    
        $$
        f(g(x)) = 
        \begin{cases}
            \begin{array}{ll}
                1 & \quad \mbox{if} \ x \leq -1 \\
                x^2 & \quad \mbox{if} \ -1 < x < 1 \\
                1 & \quad \mbox{if} \ x \geq 1
            \end{array}
        \end{cases}
        $$
        

    ```{r f-g-x, dependson = "base-plot"}
function_plot +
  stat_function(fun = function(x) f_x(g_x(x)), n = 1000) +
  ggtitle(expression(f(g(x))))
    ```

    -->
    
1. $y = g(f(x))$

    <!--

    To sketch the composite function, we first evaluate $f(x)$ for different values of $x$, and then evaluate $g(f(x))$ for different outputs of $f(x)$.
    
    * For $x \geq 1$, $f(x)$ is a constant value: 
        $$
        \begin{aligned}
        x \geq 1 \\
        f(x) = 1 \\
        g(f(x)) = f(1) = 1^2 = 1
        \end{aligned}
        $$
    
    * For $x < 1$, $f(x)$ is not constant: $f(x) = |x|$. $g(x)$ evaluates differently depending on its input, so we have two cases based on the output of $f(x)$:
    
        * if $f(x) < 2$, $g(f(x)) = f(x)^2 = |x|^2 = x^2$. This is the case when:
            $$
            \begin{aligned}
            f(x) < 2 \\
            |x| < 2 \text{ and } x < 1\\
            -2 < x < 1
            \end{aligned}
            $$
    
        * if $f(x) \geq 2$, $g(f(x)) = 4$. This is the case when:
            $$
            \begin{aligned}
            f(x) \geq 2 \\
            |x| \geq 2 \text{ and } x < 1\\
            x \leq -2 
            \end{aligned}
            $$

    * Therefore, $g(f(x))$ has the following values:
        $$
        g(f(x)) = 
        \begin{cases}
            \begin{array}{ll}
                4 & \quad \mbox{if} \ x \leq -2 \\
                x^2 & \quad \mbox{if} \ -2 < x < 1 \\
                1 & \quad \mbox{if} \ x \geq 1 
            \end{array}
        \end{cases}
        $$

    ```{r g-f-x, dependson = "base-plot"}
function_plot +
  stat_function(fun = function(x) g_x(f_x(x)), n = 1000) +
  ggtitle(expression(g(f(x))))
    ```

    -->
    
# Root finding

Find the roots (solutions) to the following quadratic equations.^[Gill 1.25]

$$x = \dfrac{-b \pm \sqrt{b^2 - 4ac}}{2a}$$
  
a. $9x^2 - 3x - 12 = 0$

    <!--

    * Factor the left-hand side.
    
        $$3 (x + 1)(3x - 4) = 0$$
    
    * Divide both sides by 3 to simplify the equation.
        
        $$(x + 1)(3x - 4) = 0$$
    
    * Find the roots of each term in the product separately by solving for $x$.
    
        $$
        \begin{aligned}
        x + 1 &= 0 \\
        x &= -1 \\
        \end{aligned} \qquad
        \begin{aligned}
        3x &= 4 \\
        x &= \dfrac{4}{3}
        \end{aligned}
        $$

    -->
    
a. $x^2 - 2x - 16 = 0$

    <!--

    1. Complete the square
    
        $$
        \begin{aligned}
        x^2 - 2x - 16 &= 0 \\
        x^2 - 2x &= 16 \\
        x^2 - 2x + 1 &= 17 \\
        (x - 1)^2 &= 17 \\
        x - 1 &= \pm \sqrt{17} \\
        x &= 1 \pm \sqrt{17}
        \end{aligned}
        $$
        
    1. Quadratic formula
        
        * Using the quadratic formula, solve for $x$
        
            $$
            \begin{aligned}
            x &= \dfrac{-(-2) \pm \sqrt{(-2)^2 - (4 \times 1 \times 16)}}{2 \times 1} \\
            x &= \dfrac{2 \pm \sqrt{4 + 64}}{2} \\
            x &= \dfrac{2 \pm \sqrt{68}}{2}
            \end{aligned}
            $$

        * Simplify the radical
        
            $$
            \begin{aligned}
            x &= \dfrac{2 \pm \sqrt{2^2 \times 17}}{2} \\
            x &= \dfrac{2 \pm 2\sqrt{17}}{2}
            \end{aligned}
            $$
        
        * Factor the greatest common divisor
        
            $$x = 1 \pm \sqrt{17}$$

    -->
    
a. $6x^2 - 6x - 6 = 0$
    
    <!--

    * Divide both sides by 6 to simplify the equation.
        
        $$x^2 - x - 1 = 0$$
            
    * Using the quadratic formula, solve for $x$
        
        $$
        \begin{aligned}
        x &= \dfrac{-(-1) \pm \sqrt{(-1)^2 - (4 \times 1 \times -1)}}{2 \times 1} \\
        x &= \dfrac{1 \pm \sqrt{1 - 4 (-1)}}{2} \\
        x &= \dfrac{1 \pm \sqrt{1 + 4}}{2} \\
        x &= \dfrac{1 \pm \sqrt{5}}{2}
        \end{aligned}
        $$
    
    -->
    

# Systems of linear equations

Solve the following systems of equations for their unknown values. If there is no solution, indicate as such.

a. Two unknowns^[OpenStax Algebra ex 7.1.12]

    $$
    \begin{aligned}
    3x - 2y &= 18 \\
    5x + 10y &= -10
    \end{aligned}
    $$

    <!--
    
    Via substitution:
    
    1. Solve for $x$ in the first equation
    
    $$
    \begin{aligned}
    3x - 2y &= 18 \\
    3x &= 18 + 2y \\
    x &= \frac{2}{3}y + 6
    \end{aligned}
    $$
    
    1. Substitute $x = \frac{2}{3}y + 6$ into the second equation and solve for $y$
    
    $$
    \begin{aligned}
    5\left( \frac{2}{3}y + 6 \right) + 10y &= -10 \\
    \frac{10}{3}y + 30 + 10y &= -10 \\
    \frac{40}{3}y + 30 &= -10 \\
    \frac{40}{3}y &= -40 \\
    y &= -3
    \end{aligned}
    $$
    
    1. Substitute $y$ back into the first equation
    
    $$
    \begin{aligned}
    3x - 2(-3) &= 18 \\
    3x + 6 &= 18 \\
    3x &= 12 \\
    x &= 4
    \end{aligned}
    $$
    
    $$x = 4, y = -3$$
    -->

a. Three unknowns^[OpenStax Algebra ex 7.2.12]

    $$
    \begin{aligned}
    5x - 2y + 3z &= 20 \\
    2x - 4y - 3z &= -9 \\
    x + 6y - 8z &= 21
    \end{aligned}
    $$
    
    <!--
    
    1. Subtract 2/5 × (equation 1) from equation 2:
    
        $$
        \begin{aligned}
        5 x - 2 y + 3 z &= 20 \\
        0 x - \frac{16 y}{5} - \frac{21 z}{5} &= -17 \\
        x + 6 y - 8 z &= 21
        \end{aligned}
        $$
    
    1. Multiply equation 2 by -5:
    
        $$
        \begin{aligned}
        5 x - 2 y + 3 z &= 20 \\
        0 x+16 y + 21 z &= 85 \\
        x + 6 y - 8 z &= 21
        \end{aligned}
        $$
    1. Subtract 1/5 × (equation 1) from equation 3:
    
        $$
        \begin{aligned}
        5 x - 2 y + 3 z &= 20 \\
        0 x+16 y + 21 z &= 85 \\
        0 x + \frac{32 y}{5} - \frac{43 z}{5} &= 17
        \end{aligned}
        $$
    1. Multiply equation 3 by 5:
    
        $$
        \begin{aligned}
        5 x - 2 y + 3 z &= 20 \\
        0 x+16 y + 21 z &= 85 \\
        0 x + 32y - 43z &= 85
        \end{aligned}
        $$
    
    1. Swap equation 2 with equation 3:
    
        $$
        \begin{aligned}
        5 x - 2 y + 3 z &= 20 \\
        0 x + 32y - 43z &= 85 \\
        0 x+16 y + 21 z &= 85 \\
        \end{aligned}
        $$
    
    1. Subtract 1/2 × (equation 2) from equation 3:
    
        $$
        \begin{aligned}
        5 x - 2 y + 3 z &= 20 \\
        0 x + 32y - 43z &= 85 \\
        0 x + 0 y + \frac{85z}{2} &= \frac{85}{2} \\
        \end{aligned}
        $$
    
    1. Multiply equation 3 by 2/85:
    
        $$
        \begin{aligned}
        5 x - 2 y + 3 z &= 20 \\
        0 x + 32y - 43z &= 85 \\
        0 x + 0 y + z &= 1 \\
        \end{aligned}
        $$
    
    1. Add 43 × (equation 3) to equation 2:
    
        $$
        \begin{aligned}
        5 x - 2 y + 3 z &= 20 \\
        0 x + 32y + 0z &= 128 \\
        0 x + 0 y + z &= 1 \\
        \end{aligned}
        $$
    
    1. Divide equation 2 by 32:
    
        $$
        \begin{aligned}
        5 x - 2 y + 3 z &= 20 \\
        0 x + y + 0z &= 4 \\
        0 x + 0 y + z &= 1 \\
        \end{aligned}
        $$
    
    1. Add 2 × (equation 2) to equation 1:
    
        $$
        \begin{aligned}
        5 x + 0y + 3 z &= 28 \\
        0 x + y + 0z &= 4 \\
        0 x + 0 y + z &= 1 \\
        \end{aligned}
        $$
    
    1. Subtract 3 × (equation 3) from equation 1:
    
        $$
        \begin{aligned}
        5 x + 0y + 0 z &= 25 \\
        0 x + y + 0z &= 4 \\
        0 x + 0 y + z &= 1 \\
        \end{aligned}
        $$
    
    1. Divide equation 1 by 5:
    
        $$
        \begin{aligned}
         x + 0y + 0 z &= 5 \\
        0 x + y + 0z &= 4 \\
        0 x + 0 y + z &= 1 \\
        \end{aligned}
        $$
    
    $$x = 5, y = 4, z = 1$$
    
    -->

a. An animal shelter has a total of 350 animals comprised of cats, dogs, and rabbits. If the number of rabbits is 5 less than one-half the number of cats, and there are 20 more cats than dogs, how many of each animal are at the shelter?^[OpenStax Algebra 7.2.54]
    
    <!--
    
    * Let $x =$ number of cats
    * Let $y =$ number of dogs
    * Let $z =$ number of rabbits
    
    This gives us the system of equations
    
    $$
    \begin{aligned}
    x + y + z &= 350 \\
    z &= \frac{1}{2}x - 5 \\
    x &= 20 + y
    \end{aligned}
    $$
    
    1. Substitute $z = \frac{1}{2}x - 5$ into the first equation
    
        $$
        \begin{aligned}
        x + y + \frac{x}{2} - 5 &= 350 \\
        \frac{3}{2}x + y - 5 &= 350
        \end{aligned}
        $$
    
    1. Substitute $x = y + 20$ into the first equation
    
        $$
        \begin{aligned}
        \frac{3}{2}(y + 20) + y - 5 &= 350 \\
        \frac{3}{2}y + 30 + y - 5 &= 350 \\
        \frac{5}{2}y + 25 &= 350 \\
        \frac{5}{2}y &= 325 \\
        y &= 130
        \end{aligned}
        $$
    
    1. Substitute $y = 130$ into the third equation
    
        $$
        \begin{aligned}
        x &= y + 20 \\
        x &= 130 + 20 \\
        x &= 150
        \end{aligned}
        $$
    
    1. Substitute $x = 150$ into the second equation
    
        $$
        \begin{aligned}
        z &= \frac{1}{2}x - 5 \\
        z &= \frac{1}{2}(150) - 5 \\
        z &= 70
        \end{aligned}
        $$
    
    $$x = 150, y = 130, z = 70$$
    
    There are 150 cats, 130 dogs, and 70 rabbits.
    
    -->

# Work with sets

Using the sets

$$
\begin{aligned}
A&=\left\{2,3,7,9,13 \right\} \\
B&=\left\{ x: 4\leq x \leq 8 \  \mbox{and} \ x \ \mbox{is an integer} \right\} \\
C&=\left\{ x: 2< x < 25 \  \mbox{and} \  x \  \mbox{is prime}  \right\} \\
D&=\left\{ 1,4,9,16,25, \ldots  \right\} \\
\end{aligned}
$$

identify the following:^[Grimmer HW1.1]

1. $A\cup B$
    
    <!--

    $E = \{2,3,4,5,6,7,8,9,13\}$, combine all integers between 4 and 8 inclusive with the numbers in set $A$.

    -->

1. ($A\cup B) \cap C$

    <!--

    $F = \{3,5,7,13\}$, Since $C$ is only prime numbers greater than 2 and less than 25, we take all the prime numbers that are also included in $E$, but remember to drop out 2 since it is not included in $C$.
    
    -->
    
1. $C \cap D$
    
    <!--

    $G = \emptyset$, there are no prime numbers in $D$, so nothing is shared between $C$ and $D$.
        
    -->

# Simplify logarithms

Express each of the following as a single logarithm:^[Grimmer HW1.3]

a. $\log(x) + \log(y) - \log(z)$
    
    <!--

    * Multiplication rule of logarithms: $\log(x \times y) = \log(x) + \log(y)$
    
    * Division rule of logarithms: $\log(\dfrac{x}{y}) = \log(x) - \log(y)$
    
    * Applying the log rules, we combine logs that are added through multiplication and then combine logs that are subtracted with division.
    
    $$\log(x) + \log(y) - \log(z)$$
    
    $$\log(xy) - \log(z)$$
    
    $$\log(\dfrac{xy}{z})$$
    
    -->
    
a. $2 \log(x) + 1$

    <!--

    * Exponentiation rule of logarithms: $\log(x^y) = y \log(x)$
    
    * $\log(e) = 1$
    
    $$2 \log(x) + 1$$
    
    $$2 \log(x) + \log(e)$$
    
    $$\log(x^2) + \log(e)$$
    
    $$\log(ex^2)$$
    
    -->
    
a. $\log(x) - 2$

    <!--

    * $\log(e) = 1$
    
    $$\log(x) - 2$$
    
    $$\log(x) - 2 \log(e)$$
    
    $$\log(x) - \log(e^2)$$
    
    $$\log(\dfrac{x}{e^2})$$
    
    -->
